Changes in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2020-01-30 analysis/paper/paper.Rmd
Korea part done

The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia from the Late Pleistocene to Middle Holocene [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Localities SDG 2, which dated around c.32,000 years, contains aunal specimens,ostrich eggshell beads,and stone artefacts including blades, retouched flakes, cores, and debries along with well-preserved hearths. In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes. <!-- Levallois! I didn't pay attentioned --> The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078–13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance]. The accumulated studies on Shuidonggou site show that the blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools are likely to be locally developed at 33ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].

@@

The origins of Late Pleistocene technological innovations in Japan is suddenly appeared in differen regions of the archipelago around 38,000 years ago accaompanying with Homo sapiens found in Okinawa Island (ca. 36,000 cal BP) [@yamaoka2012use; @izuho201421]. The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [ @bae2017origin; @izuho201421]. These innovations have been interpreted as evidnce of new foraging methods such as watercraft and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early]. Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. Ishinomoto 8-ku is one of the earliets sites dated to 39,690–34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421].

<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit. Recenly, accummulated radiocarbon data show that -->

Ishinomoto 8-ku 
the age of the EUP oc- cupation falls within the interval 39,690–34,790 cal BP.the Takuma upland in Kumamoto Prefecture A total of 500 lithic specimens were recovered from the lithic concentrations. The lithic raw materials are mainly chert and andesite. The lithic assem- blage includes trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools

----

[@izuho201421; ]

 vegetation during MIS 3 (Takahara and Hayashi), large mammal extinctions in the late Pleistocene (Iwase et al.), colonization processes of the Ryukyu Islands (Kaifu et al.), marine transport of obsidian from the Kozu Island (Ikeya), introduction of new lithic tools from the Korean Peninsula (Morisaki), trap-pit hunting (Sato), Jōmon mitochondrial DNA (Adachi et al.), and the di- versification processes of the microblade technocomplex in Hokkaido (Nakazawa and Yamada)

 “circular aggregations” (dense concentrations of stone artifacts aligned in a circle), . High-quality lithic raw mate- rials, such as obsidian, “hard shale,” and sanukite, were often transported long distances and used together with locally available chert, andesite, rhyolite, and chalcedony.

high quarlity raw materials were avaliable (70 obsidian sources). the Aira-Tn (AT) tephra, dated at 30,009±189 cal BP (Smith et al. 2013) (1) the archeological components are situated below the primary AT tephra level (>30,000 cal BP), and (2) the tool kit usually (butnot always) includes trapezoids, pointed-shaped backed blades (“knife-shaped tools” in previous terminology), and edge-ground axes. the stratigraphical chronology is based on Sagamihara Upland and Mt. Ashitaka area

only a few possi- bly pre-UP sites are known, and there is ongoing, heated debate whether or not such archaic industries existed in the Japanese archipelago

----

The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology. The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. The earliest stemmped points were made on flakes. After blades appear at c. 27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section. The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'. Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time. Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [@Bae_2010; @Lee_2013; @Lee_2016]. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin]. <!-- need to expand, mention some key site & details. what is the debate?  -->

>>>>

The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia from the Late Pleistocene to Middle Holocene [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Localities SDG 2, which dated around c.32,000 years, contains aunal specimens,ostrich eggshell beads,and stone artefacts including blades, retouched flakes, cores, and debries along with well-preserved hearths. In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes. The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078–13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance]. The accumulated studies on Shuidonggou site show that the blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools are likely to be locally developed at 33ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].

@@

The origins of Late Pleistocene technological innovations in Japan is suddenly appeared in differen regions of the archipelago around 38,000 years ago accaompanying with Homo sapiens found in Okinawa Island (ca. 36,000 cal BP) [@yamaoka2012use; @izuho201421]. The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [ @bae2017origin; @izuho201421]. These innovations have been interpreted as evidnce of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early]. Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. Ishinomoto 8-ku is one of the earliets sites dated to 39,690–34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421].

<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit. Recenly, accummulated radiocarbon data show that --> 
<<!-- dp we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->>

The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology. The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. The earliest stemmped points were made on flakes. After blades appear at c. 27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section. The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'. Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time. Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [@Bae_2010; @Lee_2013; @Lee_2016]. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin]. The origin of the stemmed point is one of the most frequently devated topic related to the modern human dispersals. Bae (2010) claims two sub-stages for these new technologies based on the presence of AT tephra (presumably earlier than 25k BP). Bae’s earlier stage is started from Koreyri site dated to 27 ka. The stage is characterized by long and thick blades and relatively large prismatic cores with stemmed points. Bae’s later stage contains a microlithic industry, for example as found at Sinbuk in southwestern Korea, which dates 25,420±190 BP and 25,000±1,000 BP [@bae2010origin]. An alternative view comes from Seong (2009) and Park (2012) who claim three sub-stages by distinguishing an initial period when stemmed points appeared before blades [@seong2009emergence; #### Park 2012 ####]. The difference between the two stages of division [@bae2010origin] and the three stages [@seong2009emergence; #### Park 2012 ####] hinges on the recognition of stemmed points as a technology that is distinctive from blades. Bae stresses the combination of stemmed point and blade as tool blanks, while Seong and Park focus on the stemmed point as a new technology, having a stem part with retouch and acute-angled tip. These different perspectives on the distinctiveness of the stemmed point are at the crux of a heated debate on the nature and cause of the transition from the earlier phase to the later phase during the Late Pleistocene.
